---
title: "European Energy"
author: "Ottilie Mitchell - 20318500"
date: "26/02/2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
header-includes: \usepackage[section]{placeins}
---
## Background
These data visualisations looked at 2 datasets based on European energy, provided by Eurostat (2020). The first dataset, Energy_types looks at energy sources over the period 2016 – 2018, detailing the country, how the energy was produced (e.g. wind, conventional thermal, nuclear, hydro, wind, solar, geothermal or other) and the amount of energy produced.  The second dataset Country_totals gives the same information, however, includes whether the energy was imported, exported, total net production, energy absorbed by pumping, and energy supplied. Both datasets include an unclear variable named “Level” that was composed of two factors.

42% of energy used within Europe comes from the EU block and 55% is imported from other countries. The energy mix in the EU differs from country to country, but is mainly made up from 5 different sources: Petroleum products (36%), natural gas (21%), fossil fuels (15%), renewable energy (15%) and nuclear energy (13%) (Eurostat, 2021).

EU countries are part of the European Green Deal (EGD) which was created in 2006 and is binding on every EU state, which aims for Europe to become the first climate neutral continent by 2050 (Sikora, 2021). The EGD is a set of policy initiatives formulated by the European commission with the aim of making Europe climate neural in 2050. With methane emissions being the second biggest contributor to climate change after CO2, the initiative requires all manmade greenhouse gas emission to be removed from the atmosphere, thereby reducing the earth net climate balance to zero (Gov.uk, 2008). The goals extend to sectors including construction, biodiversity, energy, transport and food. This ambitious target requires EU countries to cut their net emission by at least 55% by 2030 from the 1990 levels. 

The EU biodiversity strategy will put Europe’s biodiversity on a path to recovery by 2030, for the benefit of the planet (European Environment Agency, 2021). The EU is a global leader in setting standard for sustainable growth. 

This data is important in understanding individual EU countries energy production as the consequences of inaction include damage to our health, environment and economy.  

 Datasets were provided by Eurostat, accesses via Github: https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-08-04 

```{r, echo=FALSE, include=FALSE}
#install.packages("tidytuesdayR")
library("tidytuesdayR")
library(tidyverse)
library(tidyr)
library(ggplot2)
library(dplyr)
library(forcats)
library(scales)
library(ggalt)
library(plotly)

#load datasets
energy_types <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-04/energy_types.csv')
country_totals <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-04/country_totals.csv')

#first look at datasets
head(energy_types)
head(country_totals)
```

```{r, echo=FALSE, include=FALSE}
#take out level 2 so its duplicated
energy_types <- energy_types %>% 
  filter(level != "Level 2") 
```


```{r, echo=FALSE, include=FALSE}
#UK is missing
energy_types %>% 
  filter(is.na(country_name))

#replace NA with United Kingdom
energy_types <- energy_types %>% 
  mutate(country_name = case_when(
    country == "UK" ~ "United Kingdom",
    TRUE ~ country_name
  ))

country_totals %>% 
  filter(is.na(country_name))
#UK missing again

#replace NA with United Kingdom
country_totals <- country_totals %>% 
  mutate(country_name = case_when(
    country == "UK" ~ "United Kingdom",
    TRUE ~ country_name
  ))
```


```{r, echo=FALSE, include=FALSE}
#combine years and data 
energy_types <- energy_types %>%
  pivot_longer(cols = starts_with("2"), #includes 2016, 2017, 2018
               names_to = "year",
               values_to = "gigawatt_hours") %>%
  mutate(year = as.integer(year))

country_totals <- country_totals %>%
  pivot_longer(cols = starts_with("2"),
               names_to = "year",
               values_to = "gigawatt_hour") %>%
  mutate(year = as.integer(year))
```



```{r, echo=FALSE, include=FALSE, warning=FALSE}
europe_totals <- energy_types %>%
  group_by(year, type) %>%
  summarise(total_power = sum(gigawatt_hours)) %>%
  ungroup() %>%
  mutate(type = fct_reorder(type, total_power, sum)) 
```




```{r, include=TRUE, echo=FALSE}

p <- europe_totals %>%
  filter(year == 2018) %>%
  mutate(type = fct_reorder(type, total_power, sum)) %>%
  ggplot(aes(total_power, type, fill = type)) +
  geom_col() +
  scale_x_continuous(labels = comma) +
  labs(title = "Total power production in 2018 in Europe",
       y = "Type",
       x = "Total energy production (Kilowatt hours)")

p
```

```{r, echo=FALSE}
#save chart
setwd("C:/Users/User/OneDrive - Harper Adams University/C7083 - Data Visulisation/Assignments")
p
jpeg(file = "europe1.jpeg")
p
dev.off()
```

```{r, echo=FALSE, include=TRUE}
data <- country_totals %>%
  filter(year == 2018) %>% #only looking at 2018
  mutate(type = fct_reorder(type, gigawatt_hour, sum)) 

data$type <- factor(data$type, levels = c("Imports", "Exports")) #only want to see imports and exports

myColors <- ifelse(levels(data$type)=="Exports" , rgb(0.1,0.1,0.7,0.5) , 
              ifelse(levels(data$type)=="Imports", rgb(0.8,0.1,0.3,0.6),
              "grey90" ) ) #assigns colours to each variable

plot(data$type, data$gigawatt_hour, col = myColors, ylab = "Energy (Gigawatt hours)", xlab = "", main = "Total Imports and Exports in 2018 across Europe")

```

```{r, echo=FALSE}
#save chart
setwd("C:/Users/User/OneDrive - Harper Adams University/C7083 - Data Visulisation/Assignments")

plot(data$type, data$gigawatt_hour, col = myColors, ylab = "Energy (Gigawatt hours)", xlab = "", main = "Total Imports and Exports in 2018 across Europe")

jpeg(file = "europe2.jpeg")

plot(data$type, data$gigawatt_hour, col = myColors, ylab = "Energy (Gigawatt hours)", xlab = "", main = "Total Imports and Exports in 2018 across Europe")

dev.off()
```

```{r, include=TRUE, echo=FALSE}
#how has energy production changed?
plot <- europe_totals %>%  
ggplot(aes(year, total_power, colour = type)) +
  geom_line() +
  geom_point(shape=21, size=2.5) +
  scale_y_continuous(labels = comma) + #scales package
  labs(x = "Year",
       y = "Total power (Gigawatt-hours)",
       title = "How energy production has change in Europe from 2016-2018",
       fill = "Energy type") +
  theme_gray()

plot <- 
  plot +
  scale_x_continuous(name = "Year", breaks = c(2016, 2017, 2018)) 
  #+
  #scale_y_continuous(name = "Total power", trans="log2", limits = c(5000,   2000000))  #transforms the y axis onto a log scale

#having a log scale significantly changed the results of the graph, so better on a linear scale

  

ggplotly(plot)
```



## Top 5 energy producers
```{r, echo=FALSE, include=FALSE}
#reimport original data
energy_types <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-04/energy_types.csv')
country_totals <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-04/country_totals.csv')
#new dataframe includeing just the top 5 producers
top_countries <- country_totals %>%
  subset(type == "Total net production") %>% #subset by total production
  slice_max(order_by = `2018`, n = 5) #only include the top 5

#convert type into a factor and reorder
energy_types$type <- factor(energy_types$type, c("Conventional thermal", "Nuclear", "Hydro", "Wind", "Solar", "Geothermal", "Other"))

energy_types$country_name <- replace_na(energy_types$country_name, "United Kingdom") 
```

```{r, echo=FALSE, include=FALSE}
#plot conventional thermal (normal energy)
p1 <- energy_types %>%
  subset((country %in% top_countries$country) & (type == "Conventional thermal")) %>% #only includes conentional thermal
  ggplot(aes(x = `2016`, xend = `2018`, y = country_name, group = country_name)) +
  geom_dumbbell(colour_x = "red",
                size_x = 2.5,
                size = 0.75,
                colour = "black",
                colour_xend = "blue",
                size_xend = 2.5) + #aesthetics of the dumbbells
  geom_point(x = 300000, y = 4, colour = "red", size = 2.5, inherit.aes = F) + #this adds points which asts as a legend
  annotate("text", label = "2016", x = 330000, y = 4) + #adds text next to the point
  geom_point(x = 300000, y = 3.8, color = "blue", size = 2.5, inherit.aes = FALSE) + #this adds points which asts as a legend
  annotate("text", label = "2018", x = 330000, y = 3.8) + #adds text next to the point
  scale_x_continuous(labels = comma) + 
  facet_wrap(~type, ncol = 1) +
  labs(x = "Energy produced", y = element_blank())

p1

```



```{r, echo=FALSE, include=TRUE}
p2 <- energy_types %>%
  subset((country %in% top_countries$country) & (type != "Conventional thermal") & (type != "Other")) %>% #include all energy types except conventional thermal
  ggplot(aes(x = `2016`, xend = `2018`, y = country_name, group = country_name)) +
  geom_dumbbell(colour_x = "red",
                size_x = 2.5,
                size = 0.75,
                colour = "black",
                colour_xend = "blue",
                size_xend = 2.5) + #dumbbell aesthetics 
    geom_point(x = 300000, y = 4, colour = "red", size = 2.5, inherit.aes = F) + #acts as legend
  annotate("text", label = "2016", x = 330000, y = 4) +
  geom_point(x = 300000, y = 2, color = "blue", size = 2.5, inherit.aes = FALSE) +
  annotate("text", label = "2018", x = 330000, y = 2) +
  scale_x_continuous(labels = comma) +
  facet_wrap(~type, ncol = 1) + #splits the types onto layered graphs
  labs(x = "Energy produced", y = element_blank(), subtitle = "Renewable Energy produced in 2016 compared to 2018", title = "Top 5 energy producing countries in Europe")


p2
```




```{r, echo=FALSE}
#save chart
setwd("C:/Users/User/OneDrive - Harper Adams University/C7083 - Data Visulisation/Assignments")
p2
jpeg(file = "europe3.jpeg")
p2
dev.off()
```




























