---
title: "R Notebook"
output: html_notebook
---
```{r}
#install.packages("tidytuesdayR")
library("tidytuesdayR")
library(tidyverse)
library(tidyr)
library(ggplot2)
library(dplyr)
library(forcats)
library(scales)
library(ggalt)
library(plotly)

energy_types <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-04/energy_types.csv')
country_totals <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-04/country_totals.csv')

head(energy_types)
head(country_totals)
```
First look at data
*can do a map as includes country codes
*graph of energy consumption bt type over time


Looking at our country_totals data set, it looks like we have a summary of a country’s energy export/import. Where as the energy_types data set contains a breakdown of a country actual energy use by type. Looking at the levels vector, we have a Level 2 level. A tweet from Tom Mock outlines that we do not need to factor this into our data, as it is a subset of hydro. This means to include it would be to double count hydro. We can filter these entries out. Our raw energy totals are in Gigawatt Hours

```{r}
#take out level 2 so its duplicated
energy_types <- energy_types %>% 
  filter(level != "Level 2") 
```


```{r}
#replace NA with United Kingdom
energy_types %>% 
  filter(is.na(country_name))
#UK is missing
energy_types <- energy_types %>% 
  mutate(country_name = case_when(
    country == "UK" ~ "United Kingdom",
    TRUE ~ country_name
  ))

country_totals %>% 
  filter(is.na(country_name))
#UK missing again
country_totals <- country_totals %>% 
  mutate(country_name = case_when(
    country == "UK" ~ "United Kingdom",
    TRUE ~ country_name
  ))
```
##Top 5 energy producers

```{r}
top_countries <- country_totals %>%
  subset(type == "Total net production") %>%
  slice_max(order_by = `2018`, n = 5)

energy_types$type <- factor(energy_types$type, c("Conventional thermal", "Nuclear", "Hydro", "Wind", "Solar", "Geothermal", "Other"))

energy_types$country_name <- replace_na(energy_types$country_name, "United Kingdom") 
```

```{r}
p1 <- energy_types %>%
  subset((country %in% top_countries$country) & (type == "Conventional thermal")) %>%
  ggplot(aes(x = `2016`, xend = `2018`, y = country_name, group = country_name)) +
  geom_dumbbell(colour_x = "red",
                size_x = 2.5,
                size = 0.75,
                colour = "black",
                colour_xend = "blue",
                size_xend = 2.5) +
  geom_point(x = 300000, y = 4, colour = "red", size = 2.5, inherit.aes = F) +
  annotate("text", label = "2016", x = 330000, y = 4) +
  geom_point(x = 300000, y = 3.8, color = "blue", size = 2.5, inherit.aes = FALSE) +
  annotate("text", label = "2018", x = 330000, y = 3.8) +
  scale_x_continuous(labels = comma) +
  facet_wrap(~type, ncol = 1) +
  labs(x = "Energy produced", y = element_blank())

p1
```


```{r}
top_countries1 <- energy_types %>%
  subset((country %in% top_countries$country) & (type == "Conventional thermal"))

top_countries1 <- subset(top_countries1, select = -c(`2017`))

fig1 <- plot_ly(top_countries1, color = I("gray80"))

fig1 <- fig1 %>% add_segments(x = ~top_countries1$`2016`, xend = ~top_countries$`2018`, y = ~top_countries1$country_name, yend = ~top_countries1$country_name, showlegend = F)

fig1 <- fig1 %>% add_markers(x = ~top_countries$`2016`, y = ~top_countries$country_name, name = "2016", colour = I("pink"))

fig1 <- fig1 %>% add_markers(x = ~top_countries$`2016`, y = ~top_countries$country_name, name = "2018", colour = I("blue"))

fig1 <- fig1 %>% layout(
    title = "Top energy producing countries using Conventional thermal",
    xaxis = list(title = "Energy (KWh"),
    yaxis = element_blank(),
    margin = list(l = 65)
  )


fig1
```




```{r}
p2 <- energy_types %>%
  subset((country %in% top_countries$country) & (type != "Conventional thermal") & (type != "Other")) %>%
  ggplot(aes(x = `2016`, xend = `2018`, y = country_name, group = country_name)) +
  geom_dumbbell(colour_x = "red",
                size_x = 2.5,
                size = 0.75,
                colour = "black",
                colour_xend = "blue",
                size_xend = 2.5) +
  scale_x_continuous(labels = comma) +
  facet_wrap(~type, ncol = 1) +
  labs(x = "Energy produced", y = element_blank())


p2
```


```{r}
#combine years and data 
energy_types <- energy_types %>%
  pivot_longer(cols = starts_with("2"), #includes 2016, 2017, 2018
               names_to = "year",
               values_to = "gigawatt_hours") %>%
  mutate(year = as.integer(year))

country_totals <- country_totals %>%
  pivot_longer(cols = starts_with("2"),
               names_to = "year",
               values_to = "gigawatt_hour") %>%
  mutate(year = as.integer(year)) 
```

##How is power produced in Europe in 2018


```{r}
data <- country_totals %>%
  filter(year == 2018) %>%
  mutate(type = fct_reorder(type, gigawatt_hour, sum))

plot(data$type, data$gigawatt_hour, xlab = "Type", ylab = "Energy produced (Gigawatt hours)")

data1 <- energy_types %>%
  filter(year == 2018)

plot(data$type, data1$gigawatt_hours, xlab = "Type", ylab = "Energy produced (Gigawatt hours)")
```


## Total energy consumption in Europe
```{r}
europe_totals <- energy_types %>%
  group_by(year, type) %>%
  summarise(total_power = sum(gigawatt_hours)) %>%
  ungroup() %>%
  mutate(type = fct_reorder(type, total_power, sum)) 

europe_totals %>%  
ggplot(aes(year, total_power, fill = type)) +
  geom_col() +
  scale_y_continuous(labels = comma) + #scales package
  labs(x = "Year",
       y = "Total power (Gigawatt-hours)",
       title = "How energy usage has change in Europe from 2016-2018",
       fill = "Energy type")

```
```{r}
europe_totals %>%  
ggplot(aes(year, total_power, fill = type)) +
  geom_line() +
  geom_point(shape=21, color="black", fill= "green", size=2.5)
  scale_y_continuous(labels = comma) + #scales package
  labs(x = "Year",
       y = "Total power (Gigawatt-hours)",
       title = "How energy usage has change in Europe from 2016-2018",
       fill = "Energy type")
```

```{r}
plot(europe_totals$total_power~europe_totals$year)
lines(europe_totals$type~europe_totals$year)

```

```{r}
europe_totals %>%
  filter(year == 2018) %>%
  mutate(type = fct_reorder(type, total_power, sum)) %>%
  ggplot(aes(total_power, type)) +
  geom_col() +
  scale_x_continuous(labels = comma) +
  labs(title = "Total power production in 2018 in Europe",
       y = "Type",
       x = "Total energy production (Kilowatt hours)")
  
```

##Which country make up the most wind power


















